Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-10-06 manuscript/manuscript.qmd
Add new analyses and sanity checks

targets::tar_load(c(participants, stimuli, vocabulary,
					attrition_trials, attrition_participants,
					attrition_trials_vtarget, attrition_participants_vtarget,
					attrition_trials_vnone, attrition_participants_vnone,
					attrition_trials_noeach, attrition_participants_noeach,
					bvq_data, gaze_aoi, 
					data_time_related, data_time_cognate,
					data_time_related_vtarget, data_time_cognate_vtarget,
					data_time_related_vnone, data_time_cognate_vnone,
					data_time_related_noeach, data_time_cognate_noeach))

@@

Implicit naming as a paradigm to study cross-language activation. (Mani & Plunkett, 2010, 2011a).
-->

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub ([https://github.com/gongcastro/cognate-priming](https://github.com/gongcastro/cognate-priming)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I. Before every testing session, caretakers were asked to read and sign an informed consent form, and were given a small gift at the end of it

@@

We used `r n_words` distinct words included in the BVQ to create the stimuli lists. We created six stimuli lists: three in Catalan, and three in Spanish. Each list contained 32 trials, each involving a prime-target-distractor group. Each word played a role as either prime, *or* as target and distractor across the three lists in their corresponding language. For instance, the Catalan word *cadira* appeared as *prime* in the three lists, but never as *target* or *distractor*; the Catalan word *bici* appeared as *target* and *distractor* across the three lists, but never as a prime. Target-distractor pairings were held constant across the three lists in each language. For instance, in all Catalan lists the word *bici* was paired with the word *porta*. Target-distractor pairings were also yoked, so that each member of the same target-distractor pair appeared once as target and once as distractor in each list. For instance, the *bici*-*porta* paired appeared twice in each of the three Catalan lists: once with *bici* as target and *porta* as distractor, and once with *porta* as target and *bici* as distractor. This counterbalancing avoided participants encountering looking at the target word guided solely by that word having being named in a previous trial. Finally, prime words appeared only once in each list: each target-distractor pair was associated with a different prime word in both appearances. In each list, the same prime word was presented alongside a different target-distractor pair. For instance, the Catalan prime word *barret* was presented with the *bici*-*porta* target-distractor pair in one list, with the *bici*-*porta* pair in another list, and with *berenar*-*amanida* in the remaining list. The order of the trials was randomised across experimental session, so that each time a participant was tested, the order in which the prime-target-distractor was presented was randomised. Each participant was randomly assigned to one of the three list in the corresponding language (their dominant language, see @sec-lp), and always the same list across their experimental sessions in the case of a recurrent participant.

In 16 of the 32 trials of the same list (henceforth *related* trials), the prime and the target words were phonologically related, sharing phonological onset (at least first phoneme). In the other 16 trials (*unrelated* trials), prime and target did not share phonological onset. 8 of the 16 *related* trials included a cognate prime (*cognate* trials), and the other 8 included a non-cognate trials (*non-cognate* trials). A prime word was considered cognate if its Catalan and Spanish translation shared phonological onset. Especial attention was paid to avoiding semantic or taxonomic relationships between prime and target words, and between prime and distractor words. Target and distractor word pairs were phonologically unrelated (did not share phonological onset). Some of them shared semantic features or a taxonomic relationship. This is the case of words associated with especially salient referents such as animals or food. To avoid infants guiding their gaze to these objects based on their saliency, we paired animals and food items together.

@@

For each word, we created a picture with a typical referent.

@@

## Analysis 1

@@

Participants do not need to know the prime or target words, and do not need to look to both pictures.

```{r dataset-noeach}
n_trials <- nrow(attrition_trials_noeach)
n_trials_valid <- inner_join(attrition_participants_noeach,
							 attrition_trials_noeach) |> 
	filter(is_valid_participant) |> 
	pull(is_valid_trial) |> 
	sum()
n_participants_valid <- sum(attrition_participants_noeach$is_valid_participant)
n_exc_prime <- sum(!attrition_trials_noeach$is_valid_gaze_prime)
n_exc_test <- sum(!attrition_trials_noeach$is_valid_gaze_test)
n_exc_test_each <- sum(!attrition_trials_noeach$is_valid_gaze_test_each)
n_exc_vocab <- sum(!attrition_trials_noeach$is_valid_vocab)
n_exc_cognate <- sum(!attrition_participants_noeach$is_valid_cognate)
n_exc_noncognate <- sum(!attrition_participants_noeach$is_valid_noncognate)
n_exc_unrelated <- sum(!attrition_participants_noeach$is_valid_unrelated)

@@

>>>>

targets::tar_load(
	c(participants,
	  stimuli,
	  vocabulary,
	  attrition_trials, 
	  attrition_participants,
	  attrition_trials_vtarget,
	  attrition_participants_vtarget,
	  attrition_trials_vnone, 
	  attrition_participants_vnone,
	  attrition_trials_noeach,
	  attrition_participants_noeach,
	  bvq_data, 
	  gaze_aoi, 
	  looking_times,
	  data_time_related, 
	  data_time_cognate,
	  data_time_related_vtarget,
	  data_time_cognate_vtarget,
	  data_time_related_vnone,
	  data_time_cognate_vnone,
	  data_time_related_noeach, 
	  data_time_cognate_noeach))

@@

Implicit naming as a paradigm to study cross-language activation. (Mani & Plunkett, 2010, 2011a).

Floccia et al. (2020) tested toddlers in a cross-language priming paradigm. A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. Some critiques:

- Since participants heard the auditory label in one of the languages, priming through translation might be compromised.
- Translation equivalents (Exp. 1) are more strongly related at the semantic level with each other than other pairs of semantically related words that are not translations of each other (table-mesa are more strongly associated than table-silla).
- We used longitudinal participants, so we can examine the developmental trajectories of cross-language lateral inhibitory or excitatory connections better.
- Larger sample size.
-->

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/ckydb/)) and GitHub ([https://github.com/gongcastro/cognate-priming](https://github.com/gongcastro/cognate-priming)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I. Before every testing session, caregivers were asked to read and sign an informed consent form, and were given a token of appreciation at the end of it.

@@

We used `r n_words` distinct words included in the BVQ to create the stimuli lists. We created six stimuli lists: three in Catalan, and three in Spanish. Each list contained 32 trials, each involving a prime-target-distractor group. Each word played a role as either prime, *or* as target and distractor across the three lists in their corresponding language. For instance, the Catalan word *cadira* appeared as *prime* in the three lists, but never as *target* or *distractor*; the Catalan word *bici* appeared as *target* and *distractor* across the three lists, but never as a prime. Target-distractor pairings were held constant across the three lists in each language. For instance, in all Catalan lists the word *bici* was paired with the word *porta*. Target-distractor pairings were also yoked, so that each member of the same target-distractor pair appeared once as target and once as distractor in each list. For instance, the *bici*-*porta* paired appeared twice in each of the three Catalan lists: once with *bici* as target and *porta* as distractor, and once with *porta* as target and *bici* as distractor. This counterbalancing avoided participants encountering looking at the target word guided solely by that word having being named in a previous trial. Finally, prime words appeared only once in each list: each target-distractor pair was associated with a different prime word in both appearances. In each list, the same prime word was presented alongside a different target-distractor pair. For instance, the Catalan prime word *barret* was presented with the *bici*-*porta* target-distractor pair in one list, with the *bici*-*porta* pair in another list, and with *berenar*-*amanida* in the remaining list. The order of the trials was randomised across experimental session, so that each time a participant was tested, the order in which the prime-target-distractor was presented was randomised. Each participant was randomly assigned to one of the three lists in the corresponding language (their dominant language, see @sec-lp), and always the same list across their experimental sessions in the case of a recurrent participant.

In 16 of the 32 trials of the same list (henceforth *related* trials), the prime and the target words were phonologically related, sharing phonological onset (at least first phoneme). In the other 16 trials (*unrelated* trials), prime and target did not share phonological onset. 8 of the 16 *related* trials included a cognate prime (*cognate* trials), and the other 8 included a non-cognate trials (*non-cognate* trials). A prime word was considered cognate if its Catalan and Spanish translation shared phonological onset. Especial attention was paid to avoiding semantic or taxonomic relationships between prime and target words, and between prime and distractor words. Target and distractor word pairs were phonologically unrelated (did not share phonological onset). Some of them shared semantic features or a taxonomic relationship. This is the case of words associated with especially salient referents such as animals or food. To avoid infants guiding their gaze to these objects based on their saliency, we paired animals and food items together. The position of the target and distractor pictures (right or left) for each target-distractor pair was alternated, so that in one list the target would appear on the left, in another list it would appear on the right, and so on.

@@

For each word, we created a picture with a typical referent. To avoid competition between target and distractor pictures, semantically related target-distractor pairs were perceptually distinct [@floccia2020translation; @arias-trejo2010].

@@

We know present the results under four different trial-level inclusion criteria In Analysis 1, the child is required to understand the prime *and* the target words, on top of having to fixate both target and distractor  for at last 50 ms each during the target-distractor phase. In Analysis 2, the child is required to understand the target word, on top of having to fixate both target and distractor  for at last 50 ms each during the target-distractor phase. In Analysis 3, the child is not required to understand any word, but is still required to fixate both target and distractor for at last 50 ms each during the target-distractor phase. In Analysis 4, the child is not required to understand any word, nor to fixate at both target and distractor.

For reference:

* [@floccia2020] requires participants to know prime and target, but not to look at target *and* distractor (looking at at least one of them is enough).

The fact that in some time course analyses participants seem to be looking predominantly towards one of the objects (despite the target and distractor pictures having been presented less than 300 ms before) is a common finding in previous studies using preferential looking procedures [e.g., @floccia2020translation].

```{r fig-related-summary}
#| label: fig-related-summary
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Marginal posterior predictions of the GAMMs. (A) Mean posterior probability of target looking across the time course of the trial. Black lines and intervals indicate the psoterior mean and 95% credible intervals. Points indicate the mean probability of target looking across participants. (B) Difference in posterior probability of target looking between *related* and *unrelated* trials. The yellow rectangle indicates, in both A and B, the range of time points in which the 95% credible interval of the differences excluded zero."
tibble::lst(data_time_related,
			data_time_related_vtarget,
			data_time_related_vnone,
			data_time_related_noeach,
			data_time_related_vnoeach) |> 
	bind_rows(.id = "analysis") |>
	mutate(analysis = case_when(
		analysis=="data_time_related" ~ "Analysis 1",
		analysis=="data_time_related_vtarget" ~ "Analysis 2",
		analysis=="data_time_related_vnone" ~ "Analysis 3",
		analysis=="data_time_related_noeach" ~ "Analysis 4",
		analysis=="data_time_related_vnoeach" ~ "Analysis 5"
	)) |> 
	# summarise(.elog_mean = mean(.elog),
	# 		  n = n(),
	# 		  .elog_std = sd(.elog),
	# 		  .by = c(condition, timebin, analysis, lp)) |>
	# mutate(.elog_se = .elog_std/n) |> 
	ggplot(aes(timebin, .elog,
			   colour = condition,
			   fill = condition,
			   shape = condition)) +
	facet_grid(analysis~lp) +
	geom_hline(yintercept = 0.5, 
			   linewidth = 1/2,
			   colour = "black",
			   linetype = "dotted") +
	geom_smooth(method = lm, 
				se = FALSE, 
				formula = y ~ splines::bs(x, 4)) +
	# geom_errorbar(aes(ymin = .elog_mean - .elog_se*1.96, 
	# 				  ymax = .elog_mean + .elog_se*1.96),
	# 			  width = 0.25,
	# 			  linewidth = 3/4) + 
	# geom_line(size = 3/4) + 
	stat_summary(fun = mean, geom = "point") +
	stat_summary(fun.data = mean_se,
				 geom = "errorbar",
				 width = 0.25) +
	labs(x = "Time (ms)",
		 y = "P(Target looking)",
		 colour = "Condition",
		 fill = "Condition",
		 linetype = "Condition",
		 shape = "Condition") +
	scale_linetype_manual(values = rev(c("solid", "dashed"))) +
	# scale_shape_manual(values = c(1, 2)) +
	scale_x_continuous(labels = \(x) format((x * 1e2)+300, 
											big.mark = ",")) +
	theme(legend.title = element_blank(),
		  axis.title.x = element_blank()) 
```

```{r fig-cognate-summary}
#| label: fig-cognate-summary
#| fig-height: 10
#| fig-width: 8
#| fig-cap: "Marginal posterior predictions of the GAMMs. (A) Mean posterior probability of target looking across the time course of the trial. Black lines and intervals indicate the psoterior mean and 95% credible intervals. Points indicate the mean probability of target looking across participants. (B) Difference in posterior probability of target looking between *related* and *unrelated* trials. The yellow rectangle indicates, in both A and B, the range of time points in which the 95% credible interval of the differences excluded zero."
tibble::lst(data_time_cognate,
			data_time_cognate_vtarget,
			data_time_cognate_vnone,
			data_time_cognate_noeach,
			data_time_cognate_vnoeach) |> 
	bind_rows(.id = "analysis") |>
	mutate(analysis = case_when(
		analysis=="data_time_cognate" ~ "Analysis 1",
		analysis=="data_time_cognate_vtarget" ~ "Analysis 2",
		analysis=="data_time_cognate_vnone" ~ "Analysis 3",
		analysis=="data_time_cognate_noeach" ~ "Analysis 4",
		analysis=="data_time_cognate_vnoeach" ~ "Analysis 5"
	)) |> 
	# summarise(.elog_mean = mean(.elog),
	# 		  n = n(),
	# 		  .elog_std = sd(.elog),
	# 		  .by = c(condition, timebin, analysis, lp)) |>
	# mutate(.elog_se = .elog_std/n) |> 
	ggplot(aes(timebin, .elog,
			   colour = condition,
			   fill = condition,
			   shape = condition)) +
	facet_grid(analysis~lp) +
	geom_hline(yintercept = 0.5, 
			   linewidth = 1/2,
			   colour = "black",
			   linetype = "dotted") +
	geom_smooth(method = lm, 
				se = FALSE, 
				formula = y ~ splines::bs(x, 4)) +
	# geom_errorbar(aes(ymin = .elog_mean - .elog_se*1.96, 
	# 				  ymax = .elog_mean + .elog_se*1.96),
	# 			  width = 0.25,
	# 			  linewidth = 3/4) + 
	# geom_line(size = 3/4) + 
	stat_summary(fun = mean, geom = "point") +
	stat_summary(fun.data = mean_se,
				 geom = "errorbar",
				 width = 0.25) +
	labs(x = "Time (ms)",
		 y = "P(Target looking)",
		 colour = "Condition",
		 fill = "Condition",
		 linetype = "Condition",
		 shape = "Condition") +
	scale_linetype_manual(values = rev(c("solid", "dashed"))) +
	# scale_shape_manual(values = c(1, 2)) +
	scale_x_continuous(labels = \(x) format((x * 1e2)+300, 
											big.mark = ",")) +
	theme(legend.title = element_blank(),
		  axis.title.x = element_blank()) 
```
## Analysis 1

@@

Participants do not need to know the prime or target words, and do not need to look to both pictures. 
```{r dataset-noeach}
n_trials <- nrow(attrition_trials_noeach)
n_trials_valid <- inner_join(attrition_participants_noeach,
							 attrition_trials_noeach) |> 
	filter(is_valid_participant) |> 
	pull(is_valid_trial) |> 
	sum()
n_participants_valid <- sum(attrition_participants_noeach$is_valid_participant)
n_exc_prime <- sum(!attrition_trials_noeach$is_valid_gaze_prime)
n_exc_test <- sum(!attrition_trials_noeach$is_valid_gaze_test)
n_exc_test_each <- sum(!attrition_trials_noeach$is_valid_gaze_test_each)
n_exc_vocab <- sum(!attrition_trials_noeach$is_valid_vocab)
n_exc_cognate <- sum(!attrition_participants_noeach$is_valid_cognate)
n_exc_noncognate <- sum(!attrition_participants_noeach$is_valid_noncognate)
n_exc_unrelated <- sum(!attrition_participants_noeach$is_valid_unrelated)

@@

## Appendix A: imputing voabulary size scores

```{r fig-vocabulary-imputation}
vocabulary_tmp <- vocabulary |> 
	left_join(select(participants, id = id_db, age, age_group, filename),
			  by = join_by(filename)) |> 
	relocate(id, age) |> 
	select(-filename) |> 
	filter(is_imputed) |> 
	mutate(id = as.character(id))

bvq_data$vocabulary |> 
	inner_join(distinct(bvq_data$logs, id, age_group, age),
			   by = join_by(id, age_group)) |> 
	mutate(is_imputed = FALSE) |> 
	bind_rows(vocabulary_tmp) |> 
	pivot_longer(ends_with("_prop"),
				 names_to = "measure",
				 values_to = "prop") |> 
	drop_na(prop) |> 
	mutate(is_imputed = ifelse(is_imputed, "Imputed", "Observed"),
		   measure = factor(measure,
		   				 levels = c("total_prop",
		   				 		   "l1_prop",
		   				 		   "l2_prop",
		   				 		   "concept_prop",
		   				 		   "te_prop"),
		   				 labels = c("Total",
		   				 		   "L1",
		   				 		   "L2",
		   				 		   "Conceptual",
		   				 		   "TE"))) |> 
	ggplot(aes(age, prop, 
			   colour = is_imputed,
			   fill = is_imputed)) +
	facet_wrap(~measure) +
	geom_point(alpha = 1/4, size = 1) +
	geom_smooth(method = "glm", 
				formula = "y ~ x",
				method.args = list(family = "binomial"), 
				# se = FALSE,
				size = 1) +
	labs(x = "Age (months)",
		 y = "Vocabulary size",
		 colour = "Imputed") +
	theme(legend.position = c(1, 0),
		  legend.justification = c(1, 0),
		  legend.title = element_blank())

## Appendix B: distribution of prime and target looking times

```{fig-dist-prime}
looking_times |> 
	mutate(prime_time = cut(prime_time, 
							seq(0, 1.5, 0.1),
							labels = FALSE,
							include.lowest = TRUE)) |> 
	add_count(lp, age_group, name = "n_total") |> 
	count(lp, age_group, prime_time, n_total) |> 
	mutate(n = n / n_total) |> 
	ggplot(aes(prime_time, n)) +
	facet_grid(lp~age_group) +
	annotate(geom = "rect",
			 fill = "orange",
			 ymin = -Inf,
			 ymax = Inf,
			 xmin = 0,
			 xmax = 0.75*10,
			 alpha = 1/3,
			 colour = NA) +
	# annotate(label = "Excluded trials",
	# 		 geom = "text",
	# 		  x = (0.75/2)*10,
	# 		  y = 1)
	geom_col(fill = "black") +
	labs(x = "Prime looking time (s)",
		 y = "Proportion of trials") +
	scale_x_continuous(labels = \(x) x/10) +
	theme(panel.grid.major.y = element_line(linetype = "dotted",
											colour = "grey"))
```

```{fig-dist-target}
looking_times |> 
	mutate(target_time = cut(target_time, 
							seq(0, 2, 0.1),
							labels = FALSE,
							include.lowest = TRUE)) |> 
	add_count(lp, age_group, name = "n_total") |> 
	count(lp, age_group, target_time, n_total) |> 
	mutate(n = n / n_total) |> 
	ggplot(aes(target_time, n)) +
	facet_grid(lp~age_group) +
	# annotate(label = "Excluded trials",
	# 		 geom = "text",
	# 		  x = (0.75/2)*10,
	# 		  y = 1)
	geom_col(fill = "black") +
	labs(x = "Targets looking time (s)",
		 y = "Proportion of trials") +
	scale_x_continuous(labels = \(x) x/10) +
	theme(panel.grid.major.y = element_line(linetype = "dotted",
											colour = "grey"))
```

## Appendix C: prime and test looking times

```{r fig-prime-times}
#| label: fig-looking-times
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Looking time (s) to the prime AOI against audio duration. In longer audios, participants are expected to look longer to the target picture."
looking_times |>
	left_join(select(attrition_trials, 
					 filename, trial, is_valid_trial),
			  by = join_by(filename, trial)) |>
	filter(is_valid_trial) |> 
	ggplot(aes(duration,
			   prime_time,
			   colour = trial_type, 
			   fill = trial_type,
			   shape = trial_type)) +
	facet_grid(lp~age_group) +
	geom_point(alpha = 0.5, 
			   size = 0.5) +
	geom_smooth(method = "lm") +
	labs(x = "Audio duration (s)",
		 y = "Looking time (1.0-2.0 seconds)",
		 colour = "Prime type",
		 fill = "Prime type",
		 shape = "Prime type")
```

```{r fig-prime-times}
#| label: fig-target-times
#| fig-height: 6
#| fig-width: 8
#| fig-cap: "Looking time (s) to the prime AOI against audio duration. In longer audios, participants are expected to look longer to the target picture."
looking_times |>
	left_join(select(attrition_trials, 
					 filename, trial, is_valid_trial),
			  by = join_by(filename, trial)) |>
	filter(is_valid_trial) |> 
	ggplot(aes(duration,
			   target_time,
			   colour = trial_type, 
			   fill = trial_type,
			   shape = trial_type)) +
	facet_grid(lp~age_group) +
	geom_point(alpha = 0.5,
			   size = 0.5) +
	geom_smooth(method = "lm",
				formula = "y ~ x") +
	labs(x = "Audio duration (s)",
		 y = "Looking time (1.0-2.0 seconds)",
		 colour = "Prime type",
		 fill = "Prime type",
		 shape = "Prime type")
```
